Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2025-01-11 demystify.tex
updates

It is unlike that you can negotiate TA stipend (\autoref{sec:ta}), which is set by the university. Moreover, while RA is set by the adviser (\autoref{sec:ra}), it is often based on the department's standard, and it also would be unfair to other students if you get more (especially since you're new and have not proven yourself yet). However, your stipend is often increased every year by the university (automatically by a small amount).

For a specific start date or TA assignment (e.g., TA'ing a particular course), you can ask for it. Also, there is typically no moving allowance for PhD students. In short, standard things set by the university or department are unlikely negotiable.  However, you can ask for things such as computer equipment (\autoref{sec:buying-equipment}), books, (and maybe even furniture), and a conference travel budget.

@@

\subsection{How Applications are Evaluated}

\begin{itemize}
\item Each applicant is assigned to ~3 reviewers who have expertise in the applicant's research topics (so not only your field but also your specific area).  This means reviewers can be quite knowledgeable in your field and your work, your LoR writers and their work, and thus can evaluate your application quite well.

\item While we review applicants from various levels (e.g., undergrad, 1st year, 2nd year in PhD program), we evaluate and compare applicants \emph{within} their cohort level.  This means that we do not compare an undergrad, who is still applying to grad school, to a 2nd year PhD student who already has an adviser and published multiple top papers. This ensures that each applicant is evaluated fairly based on their level.

>>>>

Negotiating a TA stipend is unlikely, as it is determined by the university (\autoref{sec:ta}). For RA (\autoref{sec:ra}), advisers have more leeway as it is funded by their grants. However, they will likely do not negotiate RA stipend as they have to be fair to other students and also have to follow departmental standards.  Typically, the most wiggle room is funding in the summer (\autoref{sec:summer-funding}) where the prof. might be able to pay you more from their grants. Note that the university typically automatically increases these stipends each year by a small amount.

For a specific start date or TA assignment (e.g., TA'ing a particular course), you can ask for it. Also, there is typically no moving allowance for PhD students. In short, standard things set by the university or department are unlikely negotiable.  However, you can ask for things such as books and computer equipment (\autoref{sec:buying-equipment}).

@@

\subsection{How Applications are Evaluate}

\begin{itemize}
\item Each applicant is assigned to \emph{about 3 reviewers} who all seem to have expertise in the applicant's research topics. This means reviewers can be quite knowledgeable in your field and your work, your LoR writers and their work, and thus can evaluate your application quite in depth. This is different from PhD applications where adcom reviewers (\autoref{sec:adcom}) might not be in your field.

Note that if sometimes an original reviwer is no longer available then someone who is not in the field might review your application.  Thus in general you want to make your application accessible to a general audience in CS.

\item Reviewers attempt to compare applicants \emph{within their cohort level}.  This means that we do not compare an undergrad, who is still applying to grad school, to a 2nd year PhD student who already has an adviser and published multiple papers. This ensures that each applicant is evaluated fairly based on their level.

\item Unlike CS PhD applications, which focus on research potentials in which concrete evidence like papers, research experience, and LoRs are important, NSF GRFP also focuses on \emph{potential} but also on broader impacts, e.g., how you can help society, how you can mentor others, etc.
